---
title: "Modeling Default in Peer-to-Peer LEnding Services"
author: "Nigar Sultana"
date: '2020-07-31'

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Read csv files

```{r, message=FALSE}
df1 <- read.csv(file= "../Data/Loan2012_2013.csv",stringsAsFactors=FALSE)

df2 <- read.csv(file= "../Data/Loan2014.csv", stringsAsFactors=FALSE)

df3 <- read.csv(file= "../Data/Loan2015.csv", stringsAsFactors=FALSE)

dfCombined <- rbind(df1,df2,df3)

remove(df1,df2,df3) #remove previous data frames to consume less computation power

```

## 2. Data Wrangling

-   Dealing with empty cells

```{r, empty cells}

## Replacing blank cell with NA 

dfCombined[dfCombined==''] <- NA


## Remove empty column

dfCombined <- Filter(function(x)!all(is.na(x)), dfCombined) # (136 variables after deleting the empty columns)


## Drop columns using dplyr package (reasons of drop discussed in the report)

df_reduced <- select (dfCombined,-c(9,10, 17,18,19,21,22,28,38,39,44:46,49,53,55:57:59,63:73,75,
                                    
                                    76,77,90,92,115:136))

## Remove previous dataframes to consume less computation power

remove(dfCombined)


```

-   Data formatting

```{r, data formatting}

## Using mutate_at function to  convert the character variables to numeric

df_reduced <- df_reduced %>%
  
  mutate_at(vars(int_rate), funs (as.numeric(gsub(pattern = "%",replacement = "",x = df_reduced$int_rate)))) %>%
  
  mutate_at(vars(revol_util), funs(as.numeric(gsub(pattern = "%",replacement = "",x = df_reduced$revol_util))))


## Convert string date type to date format

chr_date <- c("issue_d", "last_pymnt_d", "last_credit_pull_d", "earliest_cr_line")


transform_date <- function(x){
  as.Date(paste0("01-", x), format = "%d-%b-%y")
} 


df_reduced<-
  df_reduced%>%
  mutate_at( .funs = funs(transform_date), .vars = chr_date,)

```

-   Dealing with missing values

```{r, dealing with missing values}

## Replace with median values 

df_reduced <- df_reduced %>%
  
  mutate_at(vars(funded_amnt,loan_amnt,funded_amnt_inv, annual_inc, dti, fico_range_low, 
                 
                 tot_hi_cred_lim), ~ifelse(is.na(.), median(., na.rm = TRUE), .))



### I have replaced the missing values of 46 numerical variables with zero

na_replace_with_zero <-
  c("mths_since_last_delinq", "mths_since_last_record",
    "mths_since_last_major_derog", "installment", "inq_last_6mths","open_acc",
    "pub_rec", "revol_bal", "total_acc","total_pymnt","total_pymnt_inv","total_rec_prncp",
    "total_rec_int","last_pymnt_amnt","last_fico_range_high","last_fico_range_low", "acc_now_delinq", "tot_coll_amt", 
    "acc_open_past_24mths", "bc_util", "chargeoff_within_12_mths", "delinq_amnt", "mo_sin_old_il_acct", "mo_sin_old_rev_tl_op",
    "mo_sin_rcnt_rev_tl_op","mo_sin_rcnt_tl", "mort_acc", "mths_since_recent_bc", "mths_since_recent_inq", "num_accts_ever_120_pd",
    "num_actv_bc_tl", "num_actv_rev_tl", "num_bc_sats", "num_bc_tl", "num_il_tl", "num_op_rev_tl", "num_rev_accts",
    "num_rev_tl_bal_gt_0", "num_sats", "num_tl_120dpd_2m", "num_tl_30dpd", "num_tl_90g_dpd_24m", "num_tl_op_past_12m","pct_tl_nvr_dlq",
    "percent_bc_gt_75", "pub_rec_bankruptcies", "tot_cur_bal", "total_rev_hi_lim", "avg_cur_bal", "bc_open_to_buy", "total_bal_ex_mort", 
    "total_bc_limit", "total_il_high_credit_limit", "revol_util" )


df_reduced <- 
  df_reduced %>%
  mutate_at(.vars = na_replace_with_zero , .funs = funs(replace(., is.na(.), 0)))


## replace missing value of last_credit_pull_d with median date

caseDay <- ymd("2019-06-01")

df_reduced <- df_reduced %>%
  
  mutate(last_credit_pull_d = case_when(is.na(last_credit_pull_d) ~ caseDay,
                                        
                                        TRUE ~ last_credit_pull_d))


## replace missing value of last_pymnt_d with maximum date

caseDay1 <- ymd("2020-06-01")

df_reduced <- df_reduced %>%
  
  mutate(last_pymnt_d = case_when(is.na(last_pymnt_d) ~ caseDay1,
                                  
                                  TRUE ~ last_pymnt_d))


### Delete observation of categorical variables

df_reduced <- 
  df_reduced %>% drop_na(term)


df_reduced <- 
  df_reduced %>% drop_na(loan_status) 

```

## 3. EDA

The exploratory data analysis has been done in Tableau and R. Details of EDA can be found in the EDA section of "ProjectFinalDraft" file.

## 4. Modeling

-Transform data for applying discrete time duration model

```{r, data transformation for modeling}

## Convert to Discrete time measurement ( get the duration time in  months)


df_clean  <- df_clean %>%
  
  mutate(
    duration_mnths = 
      (interval((df_clean$issue_d), (df_clean$last_pymnt_d)) %/% months(1)))

```

For simplicity of the analysis, I will select only 36 month loan data (details have been explained in the "ProjectFinalDraft" file).

```{r, loan month}


# convert the term variable from string to numeric and add the new variable as loan month

df_clean <- df_clean %>% 
  
  mutate(loan_month = as.numeric(substr (df_clean$term,0,3)))


df_clean <-  filter(df_clean, loan_month == 36 ) # filter observations for 36 month loan

```

**Person-period-data**

```{r, person-period-data}

## select 'Charged Off' , 'Default' ,'Late (31-120 days)' as 1 and 'Fully Paid' as 0 

df_clean <- df_clean %>%
  mutate(loan_outcome  = ifelse(loan_status %in% c('Charged Off' , 'Default' ,'Late (31-120 days)') , 1, 
                                
                                ifelse(loan_status == 'Fully Paid' , 0 , 'No info')))



## delete loan status and filter loan outcome with dummy variable

df_clean_loan<- df_clean %>% select(-loan_status) %>% filter(loan_outcome %in% c(1 , 0))


## Remove previous dataframes to consume less computation power

remove(df_clean)

## two added variables: loan mnth (transform from term variable) and loan_outcome (from loan statue; either fully paid or charged off)

df_clean_loan <- select (df_clean_loan,-c("term")) # delete the term column 


df_clean_loan$loan_outcome_2 <-  as.integer(df_clean_loan$loan_outcome) 


## take the duration months grater than 0 month and less than 41 months and get duration time quarterly

df_clean_loan <- df_clean_loan  %>% 
  
  
  filter(duration_mnths > 0 & duration_mnths < 41)  %>% 
  
  mutate(durtion_quarter = ceiling(duration_mnths/3))


## Delete variables which have duplicate columns. Date variables is not  included in the model. 

df_clean_loan <- select (df_clean_loan,-c("id","addr_state", "int_rate",
                                          
                                          "loan_month", "loan_outcome","last_credit_pull_d", "earliest_cr_line",
                                          "duration_mnths", "issue_d", "last_pymnt_d"))

```

\*\* Transforming the training and test data into quarterly observations \*\*

```{r, tranforming data into quarterly observations}

## person-period data matrix

train_df <- dataLong(df_clean_loan, 
                     
                     timeColumn = "durtion_quarter",   
                     censColumn = "loan_outcome_2",   
                     timeAsFactor = F) 


## Remove previous dataframes to consume less computation power

remove(df_clean_loan)

```

\*\* Models \*\*

```{r, models}

Model_A <- glm(formula = y ~ factor(timeInt) -1 ,
               family = binomial(link = "logit"),
               data = train_df)




## base model + borrowers' demographic information variables

Model_B<- glm(formula = y ~ factor(timeInt) -1  + grade +	purpose +
                
                verification_status + initial_list_status + delinq_2yrs + mths_since_last_delinq +  
                
                mths_since_last_record +  inq_last_6mths + annual_inc + mths_since_last_major_derog+ acc_open_past_24mths
              
              + mo_sin_old_il_acct + mo_sin_old_rev_tl_op + mo_sin_rcnt_tl + mths_since_recent_bc +
                
                mths_since_recent_inq + num_tl_30dpd + num_tl_90g_dpd_24m +  num_tl_op_past_12m,
              
              family = binomial(link = "logit"),
              
              data = train_df)



## funded_amnt_inv + dti + revol_util + fico_range_low + total_pymnt + last_pymnt_amnt +last_pymnt_amnt + last_fico_range_high + 

## tot_cur_bal + avg_cur_bal + bc_open_to_buy + tot_hi_cred_lim + total_bc_limit + total_il_high_credit_limit ,





Model_C <- glm(formula = y ~ factor(timeInt) -1  + grade +	purpose +
                 
                 verification_status + initial_list_status + delinq_2yrs + mths_since_last_delinq +  
                 
                 mths_since_last_record +  inq_last_6mths + annual_inc + mths_since_last_major_derog+ acc_open_past_24mths
               
               + mo_sin_old_il_acct + mo_sin_old_rev_tl_op + mo_sin_rcnt_tl + mths_since_recent_bc +
                 
                 mths_since_recent_inq + num_tl_30dpd + num_tl_90g_dpd_24m +  num_tl_op_past_12m + 
                 
                 funded_amnt_inv + dti + revol_util + fico_range_low + total_pymnt + last_pymnt_amnt +
                 
                 last_pymnt_amnt + last_fico_range_high + tot_cur_bal + avg_cur_bal + bc_open_to_buy +
                 
                 tot_hi_cred_lim + total_bc_limit + total_il_high_credit_limit ,
               
               family = binomial(link = "logit"),
               
               data = train_df)



## added variables: pub_rec, pub_rec_bankruptcies, revol_bal, total_acc, acc_now_delinq, bc_util, delinq_amnt, mort_acc

## num_accts_ever_120_pd, num_actv_bc_tl, num_actv_rev_tl, num_il_tl, percent_bc_gt_75, 



Model_D <- glm(formula = y ~ factor(timeInt) -1  + grade +	purpose +
                 
                 verification_status + initial_list_status + delinq_2yrs + mths_since_last_delinq +  
                 
                 mths_since_last_record +  inq_last_6mths + annual_inc + mths_since_last_major_derog+ acc_open_past_24mths
               
               + mo_sin_old_il_acct + mo_sin_old_rev_tl_op + mo_sin_rcnt_tl + mths_since_recent_bc +
                 
                 mths_since_recent_inq + num_tl_30dpd + num_tl_90g_dpd_24m +  num_tl_op_past_12m + 
                 
                 funded_amnt_inv + dti + revol_util + fico_range_low + total_pymnt + last_pymnt_amnt +
                 
                 last_pymnt_amnt + last_fico_range_high + tot_cur_bal + avg_cur_bal + bc_open_to_buy +
                 
                 tot_hi_cred_lim + total_bc_limit + total_il_high_credit_limit  +  pub_rec + pub_rec_bankruptcies +
                 
                 revol_bal + total_acc + acc_now_delinq + bc_util + delinq_amnt +  mort_acc + num_accts_ever_120_pd +
                 
                 num_actv_bc_tl + num_actv_rev_tl + num_il_tl + percent_bc_gt_75,
               
               family = binomial(link = "logit"),
               
               data = train_df)


```
